---
title: "XGBoost - Growth Analysis & Decision Rule"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## XGBoost - Growth Analysis & Decision Rule

The objevtive is to determine which small customers have **high growth potential** and should remain on **direct delivery routes**.

The goal is to identify small customers with **meaningful** growth who should stay on **direct delivery routes**.

```{r }
# Load saved models
# Set the correct working directory
setwd("C:/Users/anais/OneDrive/Documentos/Swire Coca Cola Capstone")

# Load the models
xgb_model_cases <- readRDS("xgb_model_delivered_cases.rds")
xgb_model_gallons <- readRDS("xgb_model_delivered_gallons.rds")

# Ensure XGBoost Model Exists
if (!exists("xgb_model_cases")) {
  stop("Error: XGBoost model for delivered_cases not found! Train the model first.")
}

# Step 1: Split the data into training and test sets (80% for training, 20% for testing)
set.seed(123)  # For reproducibility
train_index <- sample(1:nrow(merged_data_xgb), 0.8 * nrow(merged_data_xgb)) 
train_data <- merged_data_xgb[train_index, ]
test_data <- merged_data_xgb[-train_index, ]

# Step 2: Clean the test data (remove rows with missing values)
test_data_clean <- test_data %>%
  filter(complete.cases(.))  # Ensure no missing values

# Convert to matrix for prediction
test_matrix <- as.matrix(test_data_clean %>% select(-delivered_cases))  # Exclude the target variable

# Ensure test_matrix is numeric
test_matrix <- data.matrix(test_matrix)  # Converts the test_matrix to a numeric matrix

# Step 3: Check the structure of test_matrix to make sure it's numeric
if (!is.numeric(test_matrix)) {
  stop("Error: test_matrix is not numeric!")
}

# Step 4: Generate Predictions for Delivered Cases
merged_data_xgb$predicted_delivered_cases <- predict(xgb_model_cases, newdata = test_matrix)

# Check if the prediction was successful
head(merged_data_xgb$predicted_delivered_cases)

# Step 5: Calculate Growth Rate for Delivered Cases
merged_data_xgb <- merged_data_xgb %>%
  mutate(
    growth_rate_cases = ((predicted_delivered_cases - delivered_cases) / delivered_cases) * 100
  )

# Step 6: Categorize Customers Based on Growth Rate
merged_data_xgb <- merged_data_xgb %>%
  mutate(
    customer_decision = case_when(
      growth_rate_cases > 10 ~ "High Growth (Stay on Red Truck)",
      TRUE ~ "Stable/Declining (Move to ARTM)"
    )
  )

# Step 7: Select Relevant Columns for Analysis
customer_growth_analysis <- merged_data_xgb %>%
  select(customer_number, delivered_cases, predicted_delivered_cases, growth_rate_cases, customer_decision)

# Step 8: Display Growth Analysis Table
print(customer_growth_analysis)

# Step 9: Check how many customers fall into each category
table(merged_data_xgb$customer_decision)

# Step 10: Find the top-growing customers (excluding those with 0 past deliveries)
high_growth_customers <- merged_data_xgb %>%
  filter(customer_decision == "High Growth (Stay on Red Truck)") %>%
  arrange(desc(growth_rate_cases)) %>%
  select(
    customer_number, order_type, delivered_cases, delivered_gallons,
    frequent_order_type, cold_drink_channel, trade_channel, zip_code,
    median_delivery_cost, predicted_delivered_cases, customer_decision
  ) %>%
  head(10)

print(high_growth_customers)  # Display the table

```












FROM THE OTHER MARKDOWN

---------------- 


### Growth Analysis & Decision Rule

The objective is to determine which small customers have **high growth potential** and should remain on **direct delivery routes**.

The goal is to identify small customers with **meaningful** growth who should stay on **direct delivery routes**.

```{r }
### Growth Analysis & Decision Rule
# The goal is to identify small customers with **meaningful** growth who should stay on **direct delivery routes**.
# Load saved models
xgb_model_cases <- readRDS("xgb_model_delivered_cases.rds")
xgb_model_gallons <- readRDS("xgb_model_delivered_gallons.rds")

# Ensure XGBoost Model Exists
if (!exists("xgb_model_cases")) {
  stop("Error: XGBoost model for delivered_cases not found! Train the model first.")
}

# Step 1: Ensure the training and test data have the same columns in the same order
train_matrix <- model.matrix(~ ., data = train_data[, -which(names(train_data) == "delivered_cases")])
test_matrix <- model.matrix(~ ., data = test_data[, -which(names(test_data) == "delivered_cases")])

# Ensure the column names are in the same order
colnames(test_matrix) <- colnames(train_matrix)

# Step 2: Check for NaN or Inf values in the test matrix and replace them with 0
test_matrix[is.na(test_matrix)] <- 0
test_matrix[!is.finite(test_matrix)] <- 0

# Step 3: Now generate predictions with the fixed test_matrix
merged_data_xgb$predicted_delivered_cases <- predict(xgb_model_cases, newdata = test_matrix)

# Check if the prediction was successful
head(merged_data_xgb$predicted_delivered_cases)

# Step 4: Calculate Growth Rate for Delivered Cases
merged_data_xgb <- merged_data_xgb %>%
  mutate(
    growth_rate_cases = ((predicted_delivered_cases - delivered_cases) / delivered_cases) * 100
  )

# Step 5: Categorize Customers Based on Growth Rate
merged_data_xgb <- merged_data_xgb %>%
  mutate(
    customer_decision = case_when(
      growth_rate_cases > 10 ~ "High Growth (Stay on Red Truck)",
      TRUE ~ "Stable/Declining (Move to ARTM)"
    )
  )

# Step 6: Select Relevant Columns for Analysis
customer_growth_analysis <- merged_data_xgb %>%
  select(customer_number, delivered_cases, predicted_delivered_cases, growth_rate_cases, customer_decision)

# Step 7: Display Growth Analysis Table
print(customer_growth_analysis)
```

**Understanding each column**

- customer_number -->	Unique ID for each customer. Helps us track individual businesses.
- delivered_cases -->	The actual number of delivered cases in the past (historical data).
- predicted_delivered_cases	--> The number of delivered cases predicted by the XGBoost model for the future.
- growth_rate_cases	--> The percentage increase or decrease from past deliveries to predicted deliveries. -        - Formula: growth rate = (predicted delivered cases - delivered cases/delivered cases) * 100


## Interpreting Business Implications
```{r }
# Check how many customers fall into each category
table(merged_data_xgb$customer_decision)
```
Customers are only considered "High Growth" if:

- Their growth rate is above 10%.
- Their predicted increase is more than 10 cases (so it’s a meaningful increase)

Otherwise, they were labeled "Stable/Declining (Move to ARTM)"


```{r }
#Find the top-growing customers -> This will show the top 10 customers with the highest predicted growth.

# Find the top-growing customers (excluding those with 0 past deliveries)
high_growth_customers <- merged_data_xgb %>%
  filter(customer_decision == "High Growth (Stay on Red Truck)") %>%
  arrange(desc(growth_rate_cases)) %>%
  select(
    customer_number, order_type, delivered_cases, delivered_gallons,
    frequent_order_type, cold_drink_channel, trade_channel, zip_code,
    median_delivery_cost, predicted_delivered_cases, customer_decision
  ) %>%
  head(10)

print(high_growth_customers)  # Display the table
```

**Note:** A customer is considered "High Growth" if: 
- 1️⃣ Their predicted increase is meaningful → Let’s set a minimum threshold, like +10 delivered cases.
- 2️⃣ Growth rate is still above 10% 

**What Does This Mean for the Business?**

Example: If a customer previously ordered 0.0001 cases (essentially no orders) but is now predicted to order 53 cases, it suggests massive growth potential.

If we focus on these customers, we can strategically prioritize resources for direct delivery routes.

```{r }
# Find the stable/declining customers
stable_declining_customers <- merged_data_xgb %>%
  filter(customer_decision == "Stable/Declining (Move to ARTM)") %>%
  arrange(growth_rate_cases) %>%
  select(
    customer_number, order_type, delivered_cases, delivered_gallons,
    frequent_order_type, cold_drink_channel, trade_channel, zip_code,
    median_delivery_cost, predicted_delivered_cases, customer_decision
  ) %>%
  head(10)

print(stable_declining_customers)  # Display the table

```

**Interpretation**

- These customers have low or no growth in delivered cases.
- The delivered_cases column mostly shows 0 orders.
- Their predicted growth rate is low or negative, meaning they should be moved to ARTM.




##### Checking the Distribution of Delivered Cases to find if 10 cases is a meaningful threshold
```{r }
# Summary statistics for delivered cases
summary(merged_data_xgb$delivered_cases)

# Check distribution of delivered cases
merged_data_xgb %>%
  ggplot(aes(x = delivered_cases)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Delivered Cases",
       x = "Delivered Cases",
       y = "Count of Customers")


```

```{r }
quantile(merged_data_xgb$delivered_cases, probs = c(0.25, 0.5, 0.75, 0.90, 0.95, 0.99))

```

**Result**
  
Most customers order very few cases:

- The median number of delivered cases is 9.
- The mean is 48.62, which suggests that a few large orders are pulling the average up.
- The third quartile (75th percentile) is 28, meaning 75% of customers order 28 cases or fewer.
- 90% of customers order 77.25 cases or fewer.
- Only the top 1% of customers order 788 cases or more.


```{r }
# Find the average predicted increase
mean(merged_data_xgb$predicted_delivered_cases - merged_data_xgb$delivered_cases, na.rm = TRUE)
```
**Result**

- The average predicted increase in delivered cases is -4.08, meaning that, on average, deliveries are expected to decrease rather than increase.

- This suggests that growth predictions are identifying a relatively small number of businesses with high growth potential, while most customers are declining or staying stable.

**Reason behind the growth %** 

- 10% growth might not be enough – A small customer going from 1 to 1.1 cases isn’t really "high growth."
- Focus on real growth – Look at customers ordering at least 10 cases and growing by 10+ cases.


# Summary of XGBoost Findings

**1) Prediction Performance**

- XGBoost was used to predict delivered cases and delivered gallons based on customer data.
- RMSE improved after segmentation, indicating better prediction accuracy.

**2) Customer Growth Analysis**

- We identified which small customers have high growth potential and should stay on direct delivery routes (Red Truck).
- Customers with a growth rate above 10% and at least 10 additional cases were classified as "High Growth".
- Others were labeled as "Stable/Declining (Move to ARTM)".

**3)Key Insights from Customer Segments**

- The majority of customers fell into the Stable/Declining category.
- Some customers showed small predicted growth, but it wasn’t significant enough to justify staying on Red Truck.
- Most "High Growth" customers had a noticeable increase in predicted delivered cases.

**4)Business Recommendations**

- Focus on High Growth customers for direct delivery (Red Truck).
- Move Stable/Declining customers to ARTM to optimize resources.
- Consider reviewing low-volume customers with high growth rates to ensure their potential is meaningful.











SAVING THE DATA

```{r }
saveRDS(merged_data_xgb, "merged_data_xgb.rds")

# Saving Trained XGBoost Models
saveRDS(model_cases$model, "xgb_model_cases.rds")
saveRDS(model_gallons$model, "xgb_model_gallons.rds")
```

```{r }


```
```{r }


```

```{r }

```


```{r }

```
```{r }

```

```{r }

```


```{r }
# Step 11: Check the distribution of delivered cases to see if 10 cases is a meaningful threshold
summary(merged_data_xgb$delivered_cases)

# Check distribution of delivered cases
merged_data_xgb %>%
  ggplot(aes(x = delivered_cases)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Delivered Cases",
       x = "Delivered Cases",
       y = "Count of Customers")

# Summary statistics for delivered cases
quantile(merged_data_xgb$delivered_cases, probs = c(0.25, 0.5, 0.75, 0.90, 0.95, 0.99))

# Step 12: Find the average predicted increase
mean(merged_data_xgb$predicted_delivered_cases - merged_data_xgb$delivered_cases, na.rm = TRUE)
```

